---
title: "CH08P26 MANOVA"
author: "crl"
date: "January 12, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## MANOVA - Multiple Analysis of Variance

<hr> </hr>

### Read More

MANOVA is ANOVA with more than one response variable.


[Example here](https://www.statmethods.net/stats/anova.html)

[Read more]()

[Presentation]()

<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
library(mvnormtest)
```


## Get Data
We can load data from a csv file, such as:
```{r load data, echo = TRUE}
(datm = read.csv('data/ch08p26-manova.csv', header = TRUE, stringsAsFactors = FALSE))
```



## Descriptive Statistics

```{r descriptives}
datm_long <- gather(datm, responsevar, thedata, c('tensile', 'cof', 'bubble'), factor_key=TRUE)

datm_long %>% 
  group_by(responsevar) %>% 
  summarize(mean = mean(thedata),
            median = median(thedata),
            stdev = sd(thedata),
            min = min(thedata),
            max = max(thedata)
  )
```


## MANOVA

This example uses a factorial design (2 factor, 2 level) DOE.

### Fit the model

```{r fit manova, echo = TRUE}
# 2x2 Factorial MANOVA with 3 Dependent Variables. 
fitm <- manova(cbind(tensile, cof, bubble) ~ time*pressure, data = datm)
summary(fitm, test="Pillai")
# can also use "Wilks", "Hotelling-Lawley", and "Roy""
```

### Univariate ANOVA results for each response variable

```{r univariate anova results}
summary.aov(fitm)
```

## Univariate ANOVA

### ANOVA for tensile

```{r anova tensile}
fit_tensile = aov(tensile ~ time*pressure, data = datm)
summary(fit_tensile)
```

Plot ANOVA for tensile
```{r plot anova tensile}
plot(fit_tensile)
```

TukeyHSD
```{r tukey tensile}
TukeyHSD(aov(tensile ~ factor(time)*factor(pressure), data = datm)) 
```


### ANOVA for cof

```{r anova cof}
fit_cof = aov(cof ~ factor(time)*factor(pressure), data = datm)
summary(fit_cof)
```


Plot ANOVA for cof
```{r plot anova cof}
plot(fit_cof)
```

TukeyHSD
```{r tukey cof}
TukeyHSD(aov(cof ~ factor(time)*factor(pressure), data = datm)) 
```


### ANOVA for bubble

```{r anova bubble}
fit_bubble = aov(bubble ~ time*pressure, data = datm)
summary(fit_bubble)
```


Plot ANOVA for bubble
```{r plot anova bubble}
plot(fit_bubble)
```

TukeyHSD
```{r tukey bubble}
TukeyHSD(aov(bubble ~ factor(time)*factor(pressure), data = datm)) 
```


## End
