---
title: "CH08P17 Factor Analysis and PCA"
author: "crl"
date: "January 10, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Factor Analysis (FA) and Principal Components Analysis (PCA)

<hr> </hr>

### Read More

We should note that Factor Analysis and Principal Component Analysis might appear to be similar but serve different purposes.

* Factor Analysis uses manifest variables (things we observe) to understand latent variables (things that we cannot/do not observe)
* PCA uses only manifest (observed) variables

[Example here](https://www.statmethods.net/advstats/factor.html)

[Read more](https://web.stanford.edu/class/psych253/tutorials/FactorAnalysis.html)

[NIST PCA](http://www.itl.nist.gov/div898/handbook/pmc/section5/pmc55.htm)

[Presentation](https://docs.google.com/presentation/d/e/2PACX-1vTfz4FxrVYMDwhNPTBqQprhpIN1Cvi-i0joItkSk41mYi6VSFIgUY5MTSUprpRoqCm3DDqqE6yuTPF6/pub?start=false&loop=false&delayms=10000)

<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
library(corrplot)
```


## Get Data
We can load data from a csv file, such as:
```{r load data, echo = TRUE}
(dat_fa_orig = read.csv('data/ch08p19-discriminant-analysis.csv', header = TRUE, stringsAsFactors = FALSE))
```

<!-- Convert to long format -->
<!-- ```{r convert to matrix} -->
<!-- (dat_fa <- dat_fa_orig[,1:6] %>%  # drops 'newecon' column which is not needed for FA -->
<!--   gather(characteristic, score, creative, entrep, projects, hightech, venture)) -->

<!-- ``` -->

## Factor Analysis

### Initial review of data

Use correlation matrix and plot.
- [read more](https://rstudio-pubs-static.s3.amazonaws.com/240657_5157ff98e8204c358b2118fa69162e18.html)


Correlation Matrix
```{r correlation matrix}
round(cor(dat_fa_orig[,2:6]), 2)
```


Scatterplot Matrix
```{r scatterplot matrix}
corrplot(cor(dat_fa_orig[,2:6]), method = 'ellipse')
```



### Distinguish major factors (unrotated)

[Factor Analysis in R](https://www.promptcloud.com/blog/exploratory-factor-analysis-in-r/)
[PCA and FA in R](https://www.statmethods.net/advstats/factor.html)

More libraries:
```{r more libraries}
library(psych)
library(GPArotation)
library(nFactors)
library(factoextra)
```
```{r number of factors, echo = TRUE}
# Determine Number of Factors to Extract
matdat <- as.matrix(dat_fa_orig[,2:6])
ev <- eigen(cor(matdat)) # get eigenvalues
ap <- parallel(subject=nrow(matdat),var=ncol(matdat),
  rep=100,cent=.05)
nS <- nScree(x=ev$values, aparallel=ap$eigen$qevpea)
plotnScree(nS)
```


Unrotated Factor Loadings
    Same method as the Primer (unrotated) but different numbers in results.
    
```{r fa unrotated}
(fa_out <- fa(dat_fa_orig[,2:6], nfactor = 2, rotate = 'none', fm = 'minres'))
```

```{r scree plot}
scree(dat_fa_orig[,2:6])
```

### Distinguish major factors (rotate factors to improve meaning/interpretability)

```{r fa rotated}
(fa_out2 <- fa(dat_fa_orig[,2:6], nfactor = 2, rotate = 'varimax', fm = 'wls'))
```

## Principal Components Analysis
```{r pca}
(pca_out <- prcomp(dat_fa_orig[,2:6], scale = TRUE))
```

PCA Scree Plot
```{r plot pca 1}
fviz_eig(pca_out)
```

Graph of variables.
```{r plot pca 2}
fviz_pca_var(pca_out,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
             )
```


Biplot of individuals and variables
```{r pca biplot}
fviz_pca_biplot(pca_out, repel = TRUE,
                col.var = "#2E9FDF", # Variables color
                col.ind = "#696969"  # Individuals color
                )
```

PCA loadings - MATCHES QCI CSSBB Primer CH08P17
```{r pca loading}
# Helper function 
#::::::::::::::::::::::::::::::::::::::::
var_coord_func <- function(loadings, comp.sdev){
  loadings*comp.sdev
}
# Compute Coordinates
#::::::::::::::::::::::::::::::::::::::::
loadings <- pca_out$rotation
sdev <- pca_out$sdev
var.coord <- t(apply(loadings, 1, var_coord_func, sdev)) 
head(var.coord[, 1:4])
```


## End
