---
title: "CH08P59 F Test"
author: "crl"
date: "April 17, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## F Test

<hr> </hr>

### Overview

The _F test_ is used to compare two variances.  

* Recall that the Chi Square test can compare a sample variance to a known variance.  
* The F test is commonly used in ANOVA



### Read More

[F Test Wikipedia](https://en.wikipedia.org/wiki/F-test)  

[Examples here](http://www.sthda.com/english/wiki/f-test-compare-two-variances-in-r)  



<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(dplyr)
```



## Primer Example 

### Prepare the data set.

```{r, prepare data}
# no data set
# use simulated data
d_start <- c(9768.774, 9417.665, 9207.392, 11558.438, 11143.268, 9919.113, 11303.002, 10977.100, 9723.487)
d_oneyear <- c(9731.087, 9822.298, 9971.098, 10411.528, 10237.065, 9531.770, 9875.006)

n_start <- length(d_start)
n_oneyear <- length(d_oneyear)

cat('n = ', length(d_start), 'std dev start = ', round(sd(d_start),0), '\n')
cat('n = ', length(d_oneyear), 'std dev one year = ', round(sd(d_oneyear),0))
```


## F Test

### Critical value of F
```{r, F critical}
setalpha <- 0.05

fcrit <- qf(p = 1-setalpha, df1 = n_start-1, df2 = n_oneyear-1)

cat('F critical = ', round(fcrit, 3))
```



### Perform the F test
```{r, calculate F test, echo = TRUE}
setalpha = 0.05

(ftest_out <- var.test(x = d_start, y = d_oneyear))
```


Look at the structure of the R output for t-test
```{r, structure of ftest output}
str(ftest_out)
```


### Look at the outcome of the F test.  

```{r, F test outcome}
# calculated value of Student t
fcalc <- ftest_out$statistic

degf <- ftest_out$parameter

# for a one-tailed hypothesis test use alpha
# see calculations for F critical above

cat("degrees of freedom = ", degf[1], 'and', degf[2])
cat("\nF critical = ", round(fcrit, 3))
cat("\nF calculated = ", round(fcalc, 3))

if (fcrit < fcalc){
  cat("\nF calculated is greater than F critical therefore REJECT H0")
} else {
  cat("\nF calculated is less than F critical therefore FAIL TO REJECT H0")
}
```

### Plot the results.

```{r, plot student tF test results, echo = TRUE}
# create a plot showing the outcome
xq <- seq(0, qf(p = 0.999, df1 = n_start-1, df2 = n_oneyear-1), by = 0.01)  # the 'quantiles' (x values)
xp <- df(x = xq, df1 = degf[1], df2 = degf[2])  # the distribution (y values)

plot(x = xq, y = xp, type = 'l', 
     xlab = 'Quantiles', ylab = 'Distribution (probability)')
title(main = 'F Distribution')

# add right tail shaded area under normal curve for rejection region
polygon(c( xq[xq >= fcrit], fcrit ),  c(xp[xq >= fcrit], 0 ), col="red")

# plots the calculated F value based on degrees of freedom
abline(v = fcalc, lty = 2, col = 'darkgreen')

# plots the critical F value
abline(v = fcrit, lty = 2, col = 'red')

# add a legend
legend("topright", legend=c("Critical F value", "Calculated F value"),
       col=c("red", "darkgreen"), lty = 2, cex=0.6)
```





## End
