---
title: "CH08P03 Simple Linear Regression"
author: "crl"
date: "January 6, 2018"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Simple Linear Regression Example

## Load Packages
```{r load packages, echo = TRUE}
library(ggplot2)
library(car)
library(lmtest)
library(vrtest)
```

## Get Data
We can load data from a csv file, such as:
```{r load data}
(dat_slr = read.csv('data/ch08p03-simple-linear-regression.csv', header = TRUE, stringsAsFactors = FALSE))

```

# Create Model
```{r create model}
fit = lm(results ~ time, data = dat_slr)
```


# Model Results

## Model Summary
```{r model summary}
summary(fit)
```

## Model ANOVA Table
```{r model anova}
anova(fit)
```

## Regression Confidence Band
```{r plot with confbands}
ggplot(dat_slr, aes(x = time, y = results)) +
    geom_point(shape=1) +    # Use hollow circles
    geom_smooth(method=lm)   # Add linear regression line 
```

## Regression Confidence Band and Prediction Band together
```{r conf and pred bands}
temp_var <- predict(fit, interval="prediction")

new_df <- cbind(dat_slr, temp_var)

ggplot(new_df, aes(time, results))+
    geom_point() +
    geom_line(aes(y=lwr), color = "red", linetype = "dashed")+
    geom_line(aes(y=upr), color = "red", linetype = "dashed")+
    geom_smooth(method=lm, se=TRUE)

```

# Model Diagnostics

## Plot residuals vs fitted values
```{r plot resids vs fitted}
fitted.y = fit$fitted

# plot residuals vs fitted
plot(fitted.y, residuals(fit))
title("Residuals vs Fitted Values")
grid()
```


## Normality of Model Residuals

### Test normality using Shapiro-Wilks test 
```{r sw norm test}
res.shapiro = shapiro.test(residuals(fit))

print(res.shapiro)

if (res.shapiro$p.value < 0.05){
  print("Nonnormally distributed residuals")} else {
    print("Normally distributed residuals")
  }
```

### Residuals Normality Probability Plot
```{r normprob plot}
qqPlot(residuals(fit), main = "Normal Plot of Residuals")
```


## Overview of Influence Measures
```{r infl overview}
infl.fit = influence.measures(fit)

print("Summary of influential observations:")
summary(infl.fit)
```

```{r infl most}
id.infl = which(apply(infl.fit$is.inf, 1, any))

print("Most influential observations:")
id.infl
```



## Cook's Distance Diagnostics

### All Cooks Distance Values
```{r cooksd}
print("All Cook's D values")
(fit.cook = cooks.distance(fit))
```

## Noteworthy Cooks Distance Values
```{r cooksd infl}
infl.cook <- which(fit.cook > 3*mean(fit.cook))

print("Cases where Cooks D is noteworthy")
if (length(infl.cook)) {
  print("Influential Cook's D cases")
  infl.cook
} else {
  print("No influential Cooks D")
}
```



## Plot Cook's D 
```{r cooksd plot}
infl.cook = which(fit.cook > 3*mean(fit.cook))

# plot Cooks Distance
plot(fit.cook)
abline(h = c(1,3)*mean(fit.cook), col = 2)
title("Cook's Distance")
grid()

if (length(infl.cook) > 0){ text(infl.cook, fit.cook[infl.cook], rownames(dat)[infl.cook], pos = 2, xpd = TRUE) }
```


## Studentized Residuals Diagnostics

### Noteworthy Studentized Residuals
```{r studres noteworthy}
fit.studres = rstudent(fit)
infl.studres <- which(abs(fit.studres) > 3)

print("Cases where Studentized Residuals are noteworthy")
if (length(infl.studres)) {
  print("Influential studentized residual cases")
  infl.studres
} else {
  print("No influential studentized residual")
}
```



# Plot of Studentized Residuals
```{r studres plot}
plot(rstudent(fit), ylim = c(-4,4))
abline(h = c(-3,+3), col = 'red', lty = 3)
title('Studentized Residuals')
grid()


if (length(infl.studres) > 0){ text(infl.studres, fit.studres[infl.studres], rownames(dat)[infl.studres], pos = 2, xpd = TRUE) }
```


## Leverage Diagnostics

## LEVERAGE based on HAT MATRIX
```{r leverage hat matrix}
fit.hat = hatvalues(fit)
infl.hat <- which(fit.hat > 3*mean(fit.hat))

print("Cases where Leverage is noteworthy")
if (length(infl.hat)) {
  print("Influential leverage")
  infl.hat
} else {
  print("No influential leverage")
}

```


## Plot of Leverage Diagnostics
```{r leverage diag}
calcylim = max(3*mean(fit.hat), max(fit.hat))

plot(fit.hat, ylim = c(0,calcylim))
abline(h = 3*mean(fit.hat), col = 2)
title('Leverage')
grid()

if (length(infl.hat) > 0){ text(infl.hat, fit.hat[infl.hat], rownames(dat)[infl.hat], pos = 2, xpd = TRUE) }
```


## Functional Form Diagnostic

Is the mean of the residuals constant (= 0) across the domain of the model?

```{r fform diag}
# test for functional form
# conditional mean of residuals equal to zero
# using the RESET test

res.fform = resettest(fit)

if (res.fform$p.value < 0.05){
  print("Functional Form is Misspecified [E(e|X) <> 0]")} else {
    print("Functional Form is Adequate [E(e|X) = 0]")
  }
```


## Constant Variance Diagnostics

### Breusch-Pagan test for heteroskedasticity
```{r breuschpagan}
res.bp = bptest(fit)
print(paste('Breusch-Pagan test p-value = ', round(res.bp$p.value, 4)))

if (res.bp$p.value < 0.05){
  print("Residuals have NON-constant variance")} else {
    print("Residuals have constant variance")
  }
```

## Autocorrelation Diagnostics

### Box-Ljung Test
```{r boxljung}
# the Box-Ljung test only tests for specific individual lags

(res.box = Box.test(residuals(fit), lag = 1, type = "Ljung-Box"))

print("Box-Ljung Test for autocorrelation")

if (res.box$p.value < 0.05){
  print("Residuals NOT independent (autocorrelation)")} else {
    print("Residuals independent (no autocorrelation)")
  }
```

### Portmanteau Test
```{r portm test}
# the Auto.Q function uses a portmanteau test with multi-lags (10 by default)

portm.multilag = Auto.Q(y = residuals(fit), lags = length(residuals(fit))/2)

print("Portmanteau Test for autocorrelation")

if (portm.multilag$Pvalue < 0.05){
  print("Residuals NOT independent (autocorrelation)")} else {
    print("Residuals independent (no autocorrelation)")
  }
```

End
