---
title: "CH08P21 Cluster Analysis"
author: "crl"
date: "January 11, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Cluster Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
```

## Read about Cluster Analysis

[Example here](https://www.statmethods.net/advstats/cluster.html)

[Read more](https://rstudio-pubs-static.s3.amazonaws.com/33876_1d7794d9a86647ca90c4f182df93f0e8.html)

[Presentation](https://drive.google.com/open?id=1xf-QAsBf-Z6s8w93aQBQSCWcbHHs62fX)



## Get Data
We can load data from a csv file, such as:
```{r load data, echo = TRUE}
(cdat = read.csv('data/ch08p19-discriminant-analysis.csv', header = TRUE, stringsAsFactors = FALSE))
```

Scale the numeric columns of the data
```{r scale data}
(cdat_scaled <- scale(cdat[sapply(cdat, is.numeric)])) # standardize variables
```



## Cluster Analysis

### Fit discriminant function

```{r fit cluster, echo = TRUE}
# Ward Hierarchical Clustering
d <- dist(cdat_scaled, method = "euclidean") # distance matrix
fit_cl <- hclust(d, method="ward.D") 
```

```{r plot cluster}
plot(fit_cl) # display dendogram
groups <- cutree(fit_cl, k=2) # cut tree into 5 clusters
# draw dendogram with red borders around the 5 clusters 
rect.hclust(fit_cl, k=2, border="red")
```

### Assess the accuracy of the prediction

Percent correct for each category of the response variable.

```{r lda accuracy}
ct <- table(dat_da$newecon, predict(fit_lda, dat_da)$class)
diag(prop.table(ct, 1))
cat("\n Total percent correct \n")
sum(diag(prop.table(ct)))
```

### Scatter plot using discriminant dimension(s)

```{r lda discrim scatterplot}
plot(fit_lda) # fit from lda
```

### Partition Plot Graphs for LDA

```{r lda partition plots}
library(klaR)
partimat(factor(newecon) ~ creative + entrep + projects + hightech + venture, data = dat_da, method="lda", plot.matrix = TRUE)
```


### Scatterplot for 2 Group Problem

```{r scatterplot of group by explanatory}
pairs(dat_da[,c("creative", "entrep", "projects", "hightech", "venture")], pch = 22, bg = c('red','yellow'))
```


## Quadratic Discriminant Analysis

NOTE:  For this QDA example need to remove certain explanatory variables for demonstration purposes only.

### Fit discriminant function
```{r fit qda, echo = TRUE}
(fit_qda <- qda(newecon ~ entrep + projects + hightech, na.action = "na.omit", data = dat_da))
# CV = FALSE, prior=c(1,1)/2))
```


### Assess the accuracy of the prediction

Percent correct for each category of the response variable.

```{r qda accuracy}
ctq <- table(dat_da$newecon, predict(fit_qda, dat_da)$class)
diag(prop.table(ctq, 1))
cat("\n Total percent correct \n")
sum(diag(prop.table(ctq)))
```


### Partition Plot Graphs for QDA

```{r qda partition plots}
partimat(factor(newecon) ~ entrep + projects + hightech, data = dat_da, method="qda", plot.matrix = TRUE)
```




## End
