---
title: "CH08P65 Two Way ANOVA with Interaction"
author: "crl"
date: "May 29, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Two Way ANOVA with Interaction

<hr> </hr>

### Overview

This example is largely the same as "Main Effects" Two-Way ANOVA on page 65.  


From NIST  

* _Interaction:_ Occurs when the effect of one factor on a response depends on the level of another factor(s).  


### Read More

[Two Way ANOVA Wikipedia](https://en.wikipedia.org/wiki/Two-way_analysis_of_variance)  

[Statistical Interaction Wikipedia](https://en.wikipedia.org/wiki/Interaction_(statistics))

[Statistical Interaction NIST](https://www.itl.nist.gov/div898/handbook/pri/section7/pri7.htm#Interactions:)

[Two Way ANOVA NIST](https://en.wikipedia.org/wiki/Two-way_analysis_of_variance)  

[Two Way ANOVA with Interaction NIST](https://www.itl.nist.gov/div898/handbook/prc/section4/prc437.htm)

[ANOVA assumptions STHDA](http://www.sthda.com/english/wiki/one-way-anova-test-in-r#check-anova-assumptions-test-validity)  

[ANOVA assumptions Quick R](https://www.statmethods.net/stats/anovaAssumptions.html)  



### ANOVA Hypotheses

The null hypothesis H0 is that there is no difference between any of the sample means for each factor.  
The alternate hypothesis HA is that there is a difference between at least two of the sample means.  

The null hypothesis for the _interaction term_ is that it is not significant in the model.



<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
library(car)
library(lmtest)
library(vrtest)
```



## Primer Example 

### Prepare the data set.

Often we use stringsAsFactors = FALSE.  
In this case, with ANOVA, the categorical (explanatory) variables should be factors, 
therefore we use the default stringsAsFactors = TRUE.

```{r, load data}
dat <- read.csv('data/ch08p65-two-way-anova.csv')
glimpse(dat)
```

## Look at the data

```{r, boxplots}
ggplot(data = dat, aes(x = instructor, y = result, color = material)) + 
  geom_boxplot()
```

From just a boxplot it is not always easy to know if sample means are different.  
This is why we also use ANOVA.  


## Two Way (Main) ANOVA WITH Interaction Term

### Overview of ANOVA Models in R

[ANOVA model formula examples](https://www.statmethods.net/stats/anova.html)  


### Conducting the ANOVA analysis

Note that for the full factorial model we use "*" instead of "+".

```{r, oneway anova}
fit <- aov(result ~ material * instructor, data = dat)
summary(fit)
```


The p-value [Pr(>F)] for 'machine' tells us if there is a difference between at 
least two (or more) means. What it does not tell us is between which pairs of means 
the difference(s) exist.  For this we can use contrasts or post-hoc tests.  

Comparing results to the Primer:  

* SST "material"" matches the R output (872.4)  

* SST "instructor"" matches the R output (2005.6)  

* SST "INTERACTION"" matches the R output (53.8)  

* SSE matches the R output (240.0)  

_NOTE THAT SSE IS REDUCED FROM THE MAIN EFFECTS MODEL_  
_This is because of the addition of the interaction term._  
_Also note that the error degrees of freedom is lower._  

* F value for "material" matches the R output (21.8)  

* F value for "instructor" matches the R output (100.3)  

* F value for the interaction term matches the R output (1.34) 

* Degrees of Freedom match the R output (2, 1, 2, 12)  

MATERIAL  
The critical value of F with df = (2, 12) at alpha = 0.05 is 3.89  
The calculated value of F (21.8) is larger than F critical.  
Therefore we reject H0 that there is no difference between any of the sample means for material.  

INSTRUCTOR  
The critical value of F with df = (1, 12) at alpha = 0.05 is 4.75  
The calculated value of F (100.3) is larger than F critical.  
Therefore we reject H0 that there is no difference between any of the sample means for instructor. 

_INTERACTION TERM_  
The critical value of F with df = (2, 12) at alpha = 0.05 is 3.89    
The calculated value of F (26.9) is larger than F critical.  
Therefore we FAIL TO REJECT H0 that there is no difference due to the interaction term. 


### How can we find out where the differences exist?  

[Stat Notes - ANOVA Post-Hoc Tests](https://drive.google.com/file/d/1hs32aXJ_Nqc0mBeE9qv1yvcTYwwfX8Bg/view?usp=sharing)



```{r, posthoc}
TukeyHSD(fit)
```



### Plot of means and confidence intervals

```{r, means plot by material}
# Mean plots
# Plot response by group
# Add error bars: mean_se
# (other values include: mean_sd, mean_ci, median_iqr, ....)
library(ggpubr)
ggline(dat, x = "material", y = "result", 
       add = c("mean_se", "jitter"), 
       ylab = "Result", xlab = "Material",
       color = 'instructor')
```


```{r, means plot by instructor}
# Mean plots
# Plot weight by group
# Add error bars: mean_se
# (other values include: mean_sd, mean_ci, median_iqr, ....)
library(ggpubr)
ggline(dat, x = "instructor", y = "result", 
       add = c("mean_se", "jitter"), 
       ylab = "Result", xlab = "Instructor",
       color = 'material')
```



## Model Diagnostics  

Testing the assumptions behind the ANOVA method.  


### Plot fitted vs residuals
```{r, fitted vs resids}
# plot residuals vs fitted

# OLD CODE
# fitted.y <- fit$fitted
# plot(fitted.y, residuals(fit))
# title("Residuals vs Fitted Values")
# grid()

# NEW CODE
tmp_res <- residuals(fit)
tmp_stres <- rstudent(fit)

dat1 <- dat %>% 
  mutate(fit = fitted(fit)) %>% 
  mutate(res = tmp_res) %>% 
  mutate(stres = tmp_stres)

ggplot(dat1, aes(x = fit, y = res)) + 
  geom_point(aes(color = material, shape = instructor)) + 
  geom_smooth(method = 'loess', span = 0.75, color = 'red', linetype = 2) +
  ggtitle('Model Residuals')
```

### Plot residuals vs factor 1 (material)
```{r, resids vs factor 1}
ggplot(dat1, aes(x = material, y = res)) + 
  geom_point() + 
  ggtitle('Model Residuals vs Material')
```


### Plot residuals vs factor 2 (instructor)
```{r, resids vs factor 2}
ggplot(dat1, aes(x = instructor, y = res)) + 
  geom_point() + 
  ggtitle('Model Residuals vs Instructor')
```



### Normality diagnostics  

Test normality using Shapiro-Wilks test 
```{r, shapiro}
res.shapiro = shapiro.test(residuals(fit))

print(res.shapiro)

if (res.shapiro$p.value < 0.05){
  print("Nonnormally distributed residuals")} else {
    print("Normally distributed residuals")
  }
```


Normality plot of residuals  
```{r, normality plot resids}
qqPlot(residuals(fit), main = "Normal Plot of Residuals")
```



### Family of influence measures  
```{r, infl}
infl.fit = influence.measures(fit)

print(summary(infl.fit))

id.infl = which(apply(infl.fit$is.inf, 1, any))

print("Most influential observations:")
print(id.infl)
```



### Cook's Distance  
```{r, cooks D}
# find unusual Cooks Distance

fit.cook = cooks.distance(fit)

print("Influential Cooks D")
print(which(fit.cook > 3*mean(fit.cook)))
```



### Plot Cook's diagnostics  
```{r, cooks D plot}
fit.cook = cooks.distance(fit)

id.c = which(fit.cook > 3*mean(fit.cook))

# plot Cooks Distance
cymax <- 3*mean(fit.cook)
plot(fit.cook, ylim = c(0, cymax))
abline(h = c(1,3)*mean(fit.cook), col = c('blue', 'red'))
title("Cook's Distance")
grid()

if (length(id.c) > 0){ text(id.c, fit.cook[id.c], rownames(dat)[id.c], pos = 2, xpd = TRUE) }
```


### Print studentized residuals diagnostics  
```{r, student resids}
# STUDENTIZED RESIDUALS
fit.studres = rstudent(fit)

print("Noteworthy studentized residuals")
print(which(abs(fit.studres) > 3))
```



### Plot studentized residuals diagnostics  
```{r, student resids plot}
fit.studres = rstudent(fit)

id.sr = which(abs(fit.studres) > 3)

# plot studentized residuals
plot(rstudent(fit), ylim = c(-4,4))
abline(h = c(-3,+3), col = 'red', lty = 3)
title('Studentized Residuals')
grid()


if (length(id.sr) > 0){ text(id.sr, fit.studres[id.sr], rownames(dat)[id.sr], pos = 2, xpd = TRUE) }
```


### Print leverage diagnostics  
```{r, leverage}
# LEVERAGE based on HAT MATRIX
fit.hat = hatvalues(fit)

print("Noteworthy leverage values")
print(which(fit.hat > 3*mean(fit.hat)))
```


### Plot leverage diagnostics  
```{r, leverage plot}
fit.hat = hatvalues(fit)

id.h = which(fit.hat > 3*mean(fit.hat))


# plot leverage
plot(fit.hat)
abline(h = c(1,3)*mean(fit.hat), col = 2)
title('Leverage')
grid()


if (length(id.h) > 0){ text(id.h, fit.hat[id.h], rownames(dat)[id.h], pos = 2, xpd = TRUE) }
```


### Print functional form diagnostics  
```{r, functional form test}
# test for functional form
# conditional mean of residuals equal to zero
# using the RESET test

res.fform = resettest(fit)

if (res.fform$p.value < 0.05){
  print("Functional Form Misspecified [E(e|X) <> 0]")} else {
    print("Functional Form Adequate [E(e|X) = 0]")
  }
```


### Print constant variance (of residuals) diagnostics  
```{r, const var}
# test for heteroskedasticity
# using the Breusch-Pagan test

res.bp = bptest(fit)
res.bp$p.value

if (res.bp$p.value < 0.05){
  print("Residuals have NON-constant variance")} else {
    print("Residuals have constant variance")
  }
```


Levene's test for equal variances for "material"  
```{r, levene material}
leveneTest(result ~ material, data = dat)
```

Levene's test for equal variances for "instructor"  
```{r, levene instructor}
leveneTest(result ~ instructor, data = dat)
```




## End
