---
title: "CH08P54 Student t Test Two Sample"
author: "crl"
date: "April 9, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Student t Test Two Sample

<hr> </hr>

### Overview

The _Independent Samples Student's t-test_ is used to compare sample means between two groups.  

* The data in the two samples are independent.  


The _Dependent (Paired) Samples t-test_ is used to compare means when the samples are not independent.

* Pre- and Post- test results.  
* Beginning and Ending blood pressure for each person in a group.  
* In DOE terms, the paired t-test is a randomized block design.  
* The paired t-test can also be viewed as a repeated measures design.  


### Read More

[Student t Test Wikipedia](https://en.wikipedia.org/wiki/Student's_t-test)  

[Independent Samples Student t Test Wikipedia](https://en.wikipedia.org/wiki/Student's_t-test#Independent_two-sample_t-test)  

[Dependent Samples (Paired) Student t Test Wikipedia](https://en.wikipedia.org/wiki/Student's_t-test#Dependent_t-test_for_paired_samples)  

[Examples here](https://www.statmethods.net/stats/ttest.html)  

[Read more (Berkeley)](https://statistics.berkeley.edu/computing/r-t-tests)  

[Presentation](https://docs.google.com/presentation/d/e/2PACX-1vQHzYZ-PB8lu0ViBuok34ltHeuz4tihqMDNHvRccPIa6bDfZXEvzys3aoBWe7quW6wkEU07tRto4-L8/pub?start=false&loop=false&delayms=10000)  

<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(dplyr)
```



## Primer Example 

### Prepare the data set.

```{r, prepare data}
(dat <- read.csv("data/ch08p54-two-sample-t-test.csv", stringsAsFactors = FALSE))

# note that we set first column in the csv file as row names in the dataframe
```


## t-test assuming equal variances, equal sample sizes

### Perform the two-sample t-test
```{r calculate t equal, echo = TRUE}
setalpha = 0.05

(ttest_out_equal <- t.test(weight ~ machine, 
                           data = dat, 
                           alternative = "two.sided", 
                           var.equal = TRUE, 
                           conf.level = 1-setalpha))
```


Look at the structure of the R output for t-test
```{r, structure of ttest output}
str(ttest_out_equal)
```


### Look at the outcome of the t-test.  

```{r, t-test equal}
# calculated value of Student t
tcalc_eq <- ttest_out_equal$statistic

degf_equal <- ttest_out_equal$parameter

# for a two-sided hypothesis test use alpha/2
tcrit_lower_eq = round(qt(p = setalpha/2, df = degf_equal), 3)
tcrit_upper_eq = round(qt(p = (1 - setalpha/2), df = degf_equal), 3)

cat("degrees of freedom = ", degf_equal)
cat("\nt critical = ", abs(min(tcrit_lower_eq, tcrit_upper_eq)))
cat("\nt calculated = ", tcalc_eq)

if (tcalc_eq < tcrit_lower_eq | tcalc_eq > tcrit_upper_eq){
  cat("\nt calculated is greater than abs(t critical) therefore REJECT H0")
} else {
  cat("\nt calculated is less than abs(t critical) therefore FAIL TO REJECT H0")
}
```

### Plot the results.

```{r plot student t results, echo = TRUE}
# create a plot showing the outcome
xq <- seq(-6, 6, by = 0.01)  # the 'quantiles' (x values)
xp <- dt(x = xq, df = degf_equal)  # the distribution (y values)

plot(x = xq, y = xp, type = 'l', 
     xlab = 'Quantiles', ylab = 'Distribution (probability)')
title(main = 'Student t Distribution')

# add right tail shaded area under normal curve for rejection region
polygon(c( xq[xq >= tcrit_upper_eq], tcrit_upper_eq ),  c(xp[xq >= tcrit_upper_eq], 0 ), col="red")

# add left tail shaded area under normal curve for rejection region
polygon(c( xq[xq <= tcrit_lower_eq], tcrit_lower_eq ),  c(xp[xq <= tcrit_lower_eq], 0 ), col="red")

# plots the critical t value based on degrees of freedom (df = sample size - 1)
abline(v = tcalc_eq, lty = 2, col = 'darkgreen')

# plots the calculate t value
abline(v = c(tcrit_lower_eq, tcrit_upper_eq), lty = 2, col = 'red')

# add a legend
legend("topleft", legend=c("Critical t values", "Calculated t value"),
       col=c("red", "darkgreen"), lty = 2, cex=0.6)
```



## t-test assuming UNequal variances, equal sample sizes

### Perform the two-sample t-test
```{r calculate t unequal, echo = TRUE}
(ttest_out_unequal <- t.test(weight ~ machine, 
                             data = dat, 
                             alternative = "two.sided", 
                             var.equal = FALSE, 
                             conf.level = 1-setalpha))
```


### Look at the outcome of the t-test.  

```{r, t-test unequal}
# calculated value of Student t
tcalc_uneq <- ttest_out_unequal$statistic

degf_unequal <- ttest_out_unequal$parameter

# for a two-sided hypothesis test use alpha/2
tcrit_lower_uneq = round(qt(p = setalpha/2, df = degf_unequal), 3)
tcrit_upper_uneq = round(qt(p = (1 - setalpha/2), df = degf_unequal), 3)

cat("degrees of freedom = ", degf_unequal)
cat("\nt critical = ", abs(min(tcrit_lower_uneq, tcrit_upper_uneq)))
cat("\nt calculated = ", tcalc_uneq)

if (tcalc_uneq < tcrit_lower_uneq | tcalc_uneq > tcrit_upper_uneq){
  cat("\nt calculated is greater than abs(t critical) therefore REJECT H0")
} else {
  cat("\nt calculated is less than abs(t critical) therefore FAIL TO REJECT H0")
}
```


### What's the difference?

One difference is the degrees of freedom for the equal variance case vs. 
the unequal variance case.  The degrees of freedom for the unequal variance case 
(in this exercise) are lower, leading to a larger critical value for t (a 
smaller rejection region).



## Paired t-test assuming dependent samples

### Perform the paired t-test
```{r calculate t paired, echo = TRUE}
(ttest_out_paired <- t.test(weight ~ machine, 
                            data = dat, 
                            paired = TRUE, 
                            alternative = "two.sided", 
                            conf.level = 1-setalpha))
```


### Look at the outcome of the paired t-test.  

```{r, t-test paired}
# calculated value of Student t
tcalc_paired <- ttest_out_paired$statistic

degf_paired <- ttest_out_paired$parameter

# for a two-sided hypothesis test use alpha/2
tcrit_lower_paired = round(qt(p = setalpha/2, df = degf_paired), 3)
tcrit_upper_paired = round(qt(p = (1 - setalpha/2), df = degf_paired), 3)

cat("degrees of freedom = ", degf_paired)
cat("\nt critical = ", abs(min(tcrit_lower_paired, tcrit_upper_paired)))
cat("\nt calculated = ", tcalc_paired)

if (tcalc_paired < tcrit_lower_paired | tcalc_paired > tcrit_upper_paired){
  cat("\nt calculated is greater than abs(t critical) therefore REJECT H0")
} else {
  cat("\nt calculated is less than abs(t critical) therefore FAIL TO REJECT H0")
}
```



## End
