---
title: "CH08P19 Discriminant Analysis"
author: "crl"
date: "January 10, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Discriminant Analysis

<hr> </hr>

### Read More

[Example here](https://www.statmethods.net/advstats/discriminant.html)

[Read more](https://rstudio-pubs-static.s3.amazonaws.com/35817_2552e05f1d4e4db8ba87b334101a43da.html)

[Presentation](https://docs.google.com/presentation/d/e/2PACX-1vT9AwYm6slRlNRYry7PCXhHTJIP8zXTOrcBFP_YG39J2NrBZDj0Yakt_vG3ptz-LBOA2GYBO3l5Q4Ib/pub?start=false&loop=false&delayms=5000)

<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
library(corrplot)
library(MASS)
library(klaR)
library(mvnormtest)
```


## Get Data
We can load data from a csv file, such as:
```{r load data, echo = TRUE}
(dat_da = read.csv('data/ch08p19-discriminant-analysis.csv', header = TRUE, stringsAsFactors = FALSE))
```



## Linear Discriminant Analysis

### Fit discriminant function

```{r fit lda, echo = TRUE}
(fit_lda <- lda(newecon ~ creative + entrep + projects + hightech + venture, na.action = "na.omit", data = dat_da))
# CV = FALSE, prior=c(1,1)/2))
```

### Assess the accuracy of the prediction

Percent correct for each category of the response variable.

```{r lda accuracy}
ct <- table(dat_da$newecon, predict(fit_lda, dat_da)$class)
diag(prop.table(ct, 1))
cat("\n Total percent correct \n")
sum(diag(prop.table(ct)))
```

### Scatter plot using discriminant dimension(s)

```{r lda discrim scatterplot}
plot(fit_lda) # fit from lda
```

### Partition Plot Graphs for LDA

```{r lda partition plots}
library(klaR)
partimat(factor(newecon) ~ creative + entrep + projects + hightech + venture, data = dat_da, method="lda", plot.matrix = TRUE)
```


### Scatterplot for 2 Group Problem

```{r scatterplot of group by explanatory}
pairs(dat_da[,c("creative", "entrep", "projects", "hightech", "venture")], pch = 22, bg = c('red','yellow'))
```


## Quadratic Discriminant Analysis

NOTE:  For this QDA example need to remove certain explanatory variables for demonstration purposes only.

### Fit discriminant function
```{r fit qda, echo = TRUE}
(fit_qda <- qda(newecon ~ entrep + projects + hightech, na.action = "na.omit", data = dat_da))
# CV = FALSE, prior=c(1,1)/2))
```


### Assess the accuracy of the prediction

Percent correct for each category of the response variable.

```{r qda accuracy}
ctq <- table(dat_da$newecon, predict(fit_qda, dat_da)$class)
diag(prop.table(ctq, 1))
cat("\n Total percent correct \n")
sum(diag(prop.table(ctq)))
```


### Partition Plot Graphs for QDA

```{r qda partition plots}
partimat(factor(newecon) ~ entrep + projects + hightech, data = dat_da, method="qda", plot.matrix = TRUE)
```


## Model Assumptions

### Multivariate Normality

[Read about MVN](https://www.statmethods.net/stats/anovaAssumptions.html)

Multivariate normality requires:
- univariate normality of each variable
- normality of all linear combinations of variables

Univariate Normality
```{r univariate normality}
num_da <- dat_da[sapply(dat_da, is.numeric)]
lshap <- lapply(num_da, shapiro.test)
lres <- t(sapply(lshap, `[`, c("statistic","p.value")))

cat('Univariate Normality for each column (Shapiro-Wilk test)\n')
print(lres)

```

Q-Q Plots by Variable
```{r q-q plots}
# data wide to long
num_da_long <- gather(num_da, xvariable, xdata, names(num_da), factor_key=TRUE)

ggplot(num_da_long, aes(sample = xdata)) + 
  stat_qq() + 
  facet_wrap(~ xvariable, scales = 'free')
```


Histograms by Variable
```{r hist}
ggplot(num_da_long, aes(x = xdata)) + 
  geom_histogram() + 
  facet_wrap(~ xvariable, scales = 'free')
```


Multivariate Normality
```{r mvn test}
mat_da <- as.matrix(num_da)
mshapiro.test(t(mat_da))
```

If we have p x 1 multivariate normal random vector x vector
then the squared Mahalanobis distance between x and μ is going to be chi-square distributed with p degrees of freedom. We can use this fact to construct a Q-Q plot to assess multivariate normality.

### Graphical Assessment of Multivariate Normality
```{r mvnorm graph}
x <- mat_da  # n x p numeric matrix
center <- colMeans(x) # centroid
n <- nrow(x); p <- ncol(x); cov <- cov(x); 
d <- mahalanobis(x,center,cov) # distances 
qqplot(qchisq(ppoints(n),df=p),d,
  main="QQ Plot Assessing Multivariate Normality",
  ylab="Mahalanobis D2")
abline(a=0,b=1)
```


## End
