---
title: "CH08P19 Discriminant Analysis"
author: "crl"
date: "January 10, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Discriminant Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
library(corrplot)
library(MASS)
library(klaR)
```

## Read about Disciminant Analysis

[Example here](https://www.statmethods.net/advstats/discriminant.html)

[Read more](https://rstudio-pubs-static.s3.amazonaws.com/35817_2552e05f1d4e4db8ba87b334101a43da.html)

[Presentation](https://docs.google.com/presentation/d/e/2PACX-1vT9AwYm6slRlNRYry7PCXhHTJIP8zXTOrcBFP_YG39J2NrBZDj0Yakt_vG3ptz-LBOA2GYBO3l5Q4Ib/pub?start=false&loop=false&delayms=5000)



## Get Data
We can load data from a csv file, such as:
```{r load data, echo = TRUE}
(dat_da = read.csv('data/ch08p19-discriminant-analysis.csv', header = TRUE, stringsAsFactors = FALSE))
```



## Linear Discriminant Analysis

### Fit discriminant function

```{r fit lda, echo = TRUE}
(fit_lda <- lda(newecon ~ creative + entrep + projects + hightech + venture, na.action = "na.omit", data = dat_da))
# CV = FALSE, prior=c(1,1)/2))
```

### Assess the accuracy of the prediction

Percent correct for each category of the response variable.

```{r lda accuracy}
ct <- table(dat_da$newecon, predict(fit_lda, dat_da)$class)
diag(prop.table(ct, 1))
cat("\n Total percent correct \n")
sum(diag(prop.table(ct)))
```

### Scatter plot using discriminant dimension(s)

```{r lda discrim scatterplot}
plot(fit_lda) # fit from lda
```

### Partition Plot Graphs for LDA

```{r lda partition plots}
library(klaR)
partimat(factor(newecon) ~ creative + entrep + projects + hightech + venture, data = dat_da, method="lda", plot.matrix = TRUE)
```


### Scatterplot for 2 Group Problem

```{r scatterplot of group by explanatory}
pairs(dat_da[,c("creative", "entrep", "projects", "hightech", "venture")], pch = 22, bg = c('red','yellow'))
```


## Quadratic Discriminant Analysis

NOTE:  For this QDA example need to remove certain explanatory variables for demonstration purposes only.

### Fit discriminant function
```{r fit qda, echo = TRUE}
(fit_qda <- qda(newecon ~ entrep + projects + hightech, na.action = "na.omit", data = dat_da))
# CV = FALSE, prior=c(1,1)/2))
```


### Assess the accuracy of the prediction

Percent correct for each category of the response variable.

```{r qda accuracy}
ctq <- table(dat_da$newecon, predict(fit_qda, dat_da)$class)
diag(prop.table(ctq, 1))
cat("\n Total percent correct \n")
sum(diag(prop.table(ctq)))
```


### Partition Plot Graphs for QDA

```{r qda partition plots}
partimat(factor(newecon) ~ entrep + projects + hightech, data = dat_da, method="qda", plot.matrix = TRUE)
```




## End
