---
title: "CH08P50 Chi Square Test for Association"
author: "crl"
date: "April 3, 2018"
output: html_document
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# QCI Six Sigma Primer
## Chi Square Test for Association (page VIII-50)

<hr> </hr>

### Read More

The chi square test has several applications. 

* To compare a sample variance against a known population variance.  

* To compare goodness-of-fit of data to a distribution.  

* To test for association between categorical variables with categorical (count) responses.  

_This material addresses the chi square test for association._  


### More Resources

[Chi Square Test Wikipedia](https://en.wikipedia.org/wiki/Chi-squared_test)  

[Presentation](https://docs.google.com/presentation/d/18VCccqgjbSC2qPQAtl97FmRiec6GXO_t9WwSkKcoiWA/edit?usp=sharing)  

[Example](http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r)  

<hr> </hr>

# Analysis

## Load Packages
```{r load packages, echo = TRUE}
library(tidyverse)
library(gplots)
```


## Primer Example 8.18

Prepare the data set.

```{r, prepare data}
(dat_radio <- read.csv("data/ch08p51-chi-square-test-tabular.csv", row.names = 1))

# note that we set first column in the csv file as row names in the dataframe
```


### Now we perform the chi square test 

alpha = 0.05

The question asks: "Is there a difference between the X-ray inspectors in detection?"  
  
This is the **research hypothesis**.  

The **null hypothesis** would then be the 'status quo'  

* aka 'no difference', or 'no association found'


H0: There is no difference between the three inspectors.  H0: p1 = p2 = p3  
HA: At least one of the proportions is different.  H1: p1 <> p2 <> p3

The degrees of freedom = (rows - 1)(columns - 1) = (2-1)(3-1) = 2


Create a contingency table from the dataframe. 
```{r, contingency table}
(tbl_radio <- as.table(as.matrix(dat_radio)))
```

Look at the table.
```{r, plot table}
balloonplot(t(tbl_radio), main = "Detection", xlab = "", ylab = "",
            label = FALSE, show.margins = FALSE)
```

With a one-sided hypothesis, we put all of alpha into one tail, which affects
the critical value of chi square that defines the rejection region.

The critical value (one-sided) of X^2^ with (r-1)(c-1) degrees of freedom.
```{r, t critical}
hypalpha <- 0.05
degf <- (nrow(tbl_radio) - 1) * (ncol(tbl_radio) - 1)
chisqcrit <- round(qchisq(p = hypalpha, df = degf, lower.tail = FALSE), 3)
cat(paste('Critical value (right tail) of chisq =', chisqcrit, 'with degrees of freedom =', degf))
```

Look at both p-value of the chi square test.  

```{r chisq test, echo = TRUE}
(chisqcalc_out <- chisq.test(tbl_radio))
chisqcalc <- chisqcalc_out$statistic
```

We see that the p-value from the chi square test (0.2359) is greater than alpha (0.05).  

* Therefore, we FAIL TO REJECT H0 (null hypothesis).  
* There is not enough evidence to support the conclusion that there is a difference 
in the ability of inspectors to detect.  


Drawing conclusions

```{r, chisqtest conclusion}
if (chisqcalc < chisqcrit){
  cat("chi square calculated is greater than chisquare critical therefore REJECT H0")
} else {
  cat("chi square calculated is less than chi square critical therefore FAIL TO REJECT H0")
}
```

Graphical representation of t-test result

```{r plot chisq results, echo = TRUE}
# create a plot showing the outcome
xq <- seq(0, qchisq(p = 0.999, df = degf), by = 0.01)  # the 'quantiles' (x values)
xp <- dchisq(x = xq, df = degf)  # the distribution (y values)

plot(x = xq, y = xp, type = 'l', 
     xlab = 'Quantiles', ylab = 'Distribution (probability)')
title(main = 'Chi Square Distribution')

# plots the critical t value based on degrees of freedom (df = sample size - 1)
abline(v = chisqcalc, lty = 2, col = 'red')

# plots the calculate t value
abline(v = chisqcrit, lty = 2, col = 'darkgreen')

# add a legend
legend("topright", legend=c("Critical chi square values", "Calculated chi square value"),
       col=c("darkgreen", "red"), lty = 2, cex=0.6)
```





## End
